---
title: "Module 4 — Which Statistical Test?"
subtitle: "Statistical Analysis in R"
author: "Josh Lim"
date: today
format: revealjs
execute:
  echo: true
---

# Choosing the Right Test

## The Big Question

**Three questions determine your test:**

. . .

1. What is my **outcome** (dependent variable)?
   - Continuous (SBP, BMI) or binary (diabetic yes/no)?

. . .

2. What is my **predictor** (independent variable)?
   - Categorical (2 groups? 3+ groups?) or continuous?

. . .

3. What is my **study design**?
   - Independent groups or paired measurements?

---

## Decision Flowchart

```
What is your OUTCOME?
│
├── CONTINUOUS (SBP, BMI, Cholesterol)
│   ├── Categorical predictor, 2 groups → Two-sample t-test
│   ├── Categorical predictor, 3+ groups → One-way ANOVA
│   └── Continuous predictor → Simple linear regression
│
└── BINARY (Hypertensive yes/no, Diabetic yes/no)
    ├── Categorical predictor → Chi-square test
    └── Continuous or multiple predictors → Logistic regression
```

---

## Overview: All Tests We Cover Today {.smaller}

| Test | Outcome | Predictor | R Function |
|------|---------|-----------|------------|
| Two-sample t-test | Continuous | Categorical (2 groups) | `t.test()` |
| One-way ANOVA | Continuous | Categorical (3+ groups) | `aov()` |
| Paired t-test | Continuous | Paired (2 time points) | `t.test(..., paired=TRUE)` |
| Chi-square | Categorical | Categorical | `chisq.test()` |
| Linear regression | Continuous | Continuous | `lm()` |
| Logistic regression | Binary (0/1) | Any | `glm(..., family=binomial)` |
| Wilcoxon rank-sum | Continuous* | Categorical (2 groups) | `wilcox.test()` |

. . .

*Use non-parametric tests when normality assumptions are violated or samples are small.

# Two-Sample t-Test

## Two-Sample t-Test: Concept

**Research Q:** Is mean SBP different between smokers and non-smokers?

. . .

**When to use it:**

- Outcome is continuous (SBP in mmHg)
- Predictor is categorical with exactly 2 independent groups
- Groups are measured separately

. . .

**Core assumptions:**

- Observations are independent within and between groups
- Outcome is approximately normal per group (or $n > 30$)
- R uses Welch's t-test by default — relaxes equal variance

---

## Two-Sample t-Test: Null Hypothesis

**Null hypothesis:**

$$H_0: \mu_{\text{smokers}} = \mu_{\text{non-smokers}}$$

. . .

**What rejection means:**

- We have evidence the group means differ
- Does not tell us *why* they differ (correlation, not causation)

---

## Two-Sample t-Test in R

```{r}
#| echo: true
#| output-location: column-fragment
nhanes <- read.csv("../../data/nhanes_sample.csv")

# Formula interface: outcome ~ group
t.test(SBP ~ Smoker, data = nhanes)
```


. . .

- **t**: test statistic; larger |t| = more evidence against $H_0$
- **p-value**: probability of this t if $H_0$ were true
- **95% CI**: interval for the difference in means

# One-Way ANOVA

## One-Way ANOVA: Concept

**Research Q:** Does mean SBP differ across racial/ethnic groups?

. . .

**When to use it:**

- Outcome is continuous (SBP, BMI)
- Predictor is categorical with **3 or more independent groups**
- Extends the two-sample t-test to multiple groups

. . .

**Why not just run multiple t-tests?**

Running 3+ pairwise t-tests inflates the Type I error rate. ANOVA tests all groups simultaneously.

---

## One-Way ANOVA: Null Hypothesis {.smaller}

$$H_0: \mu_1 = \mu_2 = \mu_3 = \cdots = \mu_k$$

All group means are equal in the population.

. . .

**What rejection means:**

- At least one group mean differs from the others
- Does **not** tell you *which* groups differ — need follow-up tests

. . .

**Core assumptions:**

- Observations independent within and across groups
- Outcome approximately normal per group (or $n$ large)
- Roughly equal variance across groups

---

## One-Way ANOVA in R

```{r}
#| echo: true
#| output-location: column-fragment
nhanes <- read.csv("../../data/nhanes_sample.csv")

fit_aov <- aov(SBP ~ Race, data = nhanes)
summary(fit_aov)
```

. . .

- **F value**: ratio of between-group variance to within-group variance
- **Pr(>F)**: p-value — evidence against equal group means

---

## Post-Hoc Tests: Which Groups Differ? {.smaller}

If the ANOVA F-test is significant, use `TukeyHSD()` to identify which pairs differ:

```{r}
#| echo: true
#| output-location: column-fragment
TukeyHSD(fit_aov)
```

- Each row is one pairwise comparison
- **diff**: mean difference between groups
- **p adj**: p-value adjusted for multiple comparisons

---

# Paired t-Test

## Paired t-Test: Concept

**Research Q:** Did an exercise intervention lower SBP over 8 weeks?

. . .

**When to use it:**

- Outcome is continuous
- Each observation in group 1 is matched to exactly one in group 2
- Most common case: before/after on the same person

. . .

**Why pairing matters:**

- Each person is their own control — removes between-person variability
- Equivalent to a one-sample t-test on differences $d_i = \text{post}_i - \text{pre}_i$
- $H_0: \mu_d = 0$

---

## Paired t-Test: Key Warning

**Mistake to avoid:**

Using an unpaired test on paired data loses statistical power and can give wrong answers.

. . .

**Rule of thumb:**

- Same person measured twice → always use `paired = TRUE`
- Matched case-control → always use `paired = TRUE`

---

## Paired t-Test in R

```{r}
#| echo: true
#| output-location: column-fragment
pre  <- nhanes$SBP[1:50]
set.seed(42)
post <- pre - rnorm(50, mean = 5, sd = 8)

t.test(pre, post, paired = TRUE)   # correct
t.test(pre, post, paired = FALSE)  # wrong — ignores pairing
```

- Paired test gives smaller p-value
- Correctly accounts for within-person correlation

# Chi-Square Test

## Chi-Square Test: Concept

**Research Q:** Is smoking associated with diabetes?

. . .

**When to use it:**

- Both variables are categorical
- Testing whether distribution of one variable differs by the other
- Most common: two binary variables in a 2x2 contingency table

. . .

**Core assumption:** Expected cell counts $\geq 5$ in every cell

- If violated: use Fisher's Exact Test (`fisher.test()`)

---

## Chi-Square: The Statistic

**Null hypothesis:** Smoking and diabetes are independent

. . .

The chi-square statistic compares observed to expected counts:

$$\chi^2 = \sum \frac{(O - E)^2}{E}$$

. . .

- Large $\chi^2$ means observed counts are far from expected
- df = (rows - 1)(cols - 1) = 1 for a 2x2 table

---

## Chi-Square Test in R

```{r}
#| echo: true
ct <- table(nhanes$Smoker, nhanes$Diabetes)
dimnames(ct) <- list(Smoker   = c("Non-smoker", "Smoker"),
                     Diabetes = c("No", "Yes"))
ct
```

. . .

```{r}
#| echo: true
chisq.test(ct)
```

. . .

- **X-squared**: the chi-square statistic
- **p-value**: evidence against independence

# Simple Linear Regression

## Simple Linear Regression: Concept

**Research Q:** How does BMI predict systolic blood pressure?

. . .

**When to use it:**

- Outcome is continuous (SBP)
- Predictor is continuous (BMI)
- Goal: quantify the relationship and predict values

. . .

**The model:**

$$\text{SBP}_i = \beta_0 + \beta_1 \cdot \text{BMI}_i + \varepsilon_i$$

---

## Linear Regression: Model Terms

- $\beta_0$: intercept (expected SBP when BMI = 0)
- $\beta_1$: slope — SBP change per 1-unit increase in BMI
- $\varepsilon_i$: residual error

. . .

**Core assumptions (LINE):**

- **L**inearity — X-Y relationship is linear
- **I**ndependence — observations are independent
- **N**ormality — residuals are approximately normal
- **E**qual variance — residuals have constant spread

---

## Linear Regression in R

```{r}
#| echo: true
#| output-location: column-fragment
fit <- lm(SBP ~ BMI, data = nhanes)
summary(fit)
```

- **Estimate (BMI)**: mmHg change per 1 kg/m² increase in BMI
- **Pr(>|t|)**: p-value for $H_0: \beta_1 = 0$
- **R-squared**: variance in SBP explained by BMI

---

## Checking Assumptions: Residuals vs Fitted

```{r}
#| echo: true
#| fig-height: 5
#| fig-width: 6
#| output-location: column-fragment
# which=1 shows residuals vs fitted values
# Random scatter = linearity holds
plot(fit, which = 1)
```

---

## All Four Diagnostic Plots

```{r}
#| echo: true
#| fig-height: 5
#| fig-width: 9
par(mfrow = c(2, 2))
plot(fit)
par(mfrow = c(1, 1))
```

---

## What to Look For in Diagnostics {.smaller}

- **Residuals vs Fitted**: random scatter around zero — no curve
- **Normal Q-Q**: points follow the diagonal line — normal residuals
- **Scale-Location**: flat horizontal band — constant variance
- **Residuals vs Leverage**: no points beyond Cook's distance — no influential outliers

# Logistic Regression

## Logistic Regression: Concept

**Outcome is binary — exactly two values: 0 or 1.**

. . .

- `hypertensive`: 0 = no, 1 = yes
- `Diabetes`: 0 = no, 1 = yes
- Any yes/no clinical outcome

. . .

**What we are modeling:**

$$P(\text{outcome} = 1 \mid X_1, X_2, \ldots)$$

The **probability** that a person is a "1" given their predictors.

---

## Why Not Linear Regression?

**Research Q:** What factors predict hypertension (SBP $\geq$ 140)?

. . .

Linear regression on a 0/1 outcome produces nonsense:

- Predicted probabilities outside [0, 1] (e.g., −0.2 or 1.4)
- Residuals are not normal — assumption violated

. . .

**The fix:** logistic regression squashes predictions into (0, 1) using the logistic function:

$$p = \frac{e^{\beta_0 + \beta_1 X}}{1 + e^{\beta_0 + \beta_1 X}}$$

---

## The Logistic Model: Log-Odds {.smaller}

Rather than modeling $p$ directly, logistic regression models the **log-odds**:

$$\log\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \cdots$$

. . .

This is a linear equation — same structure as `lm()` — but the outcome is transformed so predictions always map to valid probabilities.

. . .

**Interpreting coefficients:**

| Coefficient | Interpretation |
|-------------|---------------|
| $\beta_1 > 0$ | Predictor increases probability of outcome = 1 |
| $\beta_1 < 0$ | Predictor decreases probability of outcome = 1 |
| $e^{\beta_1}$ | **Odds Ratio** per 1-unit increase in $X_1$ |

---

## Logistic Regression in R

```{r}
#| echo: true
#| output-location: column-fragment
nhanes$hypertensive <- ifelse(nhanes$SBP >= 140, 1, 0)

fit_log <- glm(hypertensive ~ BMI + Smoker,
               data   = nhanes,
               family = binomial)
summary(fit_log)
```

---

## Interpreting Logistic Regression Output

```{r}
#| echo: true
exp(coef(fit_log))   # convert log-odds to odds ratios
```

. . .

- **Estimate**: log-odds coefficient — exponentiate to get OR
- **Pr(>|z|)**: p-value for the Wald test
- **BMI OR**: e.g., 1.08 means 8% higher odds per 1-unit BMI increase
- **Smoker OR**: odds ratio comparing smokers to non-smokers, adjusted for BMI

. . .

OR > 1 = higher odds; OR < 1 = lower odds

# Non-Parametric Tests

## When to Go Non-Parametric

**Parametric tests assume:** approximately normal distributions, continuous measurement

. . .

**Use non-parametric alternatives when:**

- Small samples ($n < 20$ per group) and data look non-normal
- Severely skewed distributions (e.g., income, length of stay)
- Ordinal outcomes (pain scale 1-10, Likert items)
- Outliers that cannot be justified removing

. . .

**The trade-off:**

- Non-parametric: fewer assumptions, but less powerful when normality holds
- They compare ranks rather than raw values

---

## Wilcoxon Rank-Sum Test

**Non-parametric alternative to the two-sample t-test**

. . .

**How it works:**

1. Pool all observations from both groups and rank them
2. Sum the ranks within each group
3. Test whether rank sums differ more than expected by chance

. . .

**Research Q:** Is SBP distribution different between smokers and non-smokers?

---

## Wilcoxon Rank-Sum in R

```{r}
#| echo: true
wilcox.test(SBP ~ Smoker, data = nhanes)
```

. . .

- **W**: the test statistic (sum of ranks)
- **p-value**: evidence against identical distributions
- No normality assumption required

# Wrapping Up

## Assumptions Matter: A Quick Checklist {.smaller}

::: {style="font-size: 0.75em"}
| Test | Key Assumptions |
|------|-----------------|
| Two-sample t-test | Normality per group (or $n \geq 30$); independence |
| One-way ANOVA | Normality per group; roughly equal variances; independence |
| Paired t-test | Differences approximately normal; pairs matched |
| Chi-square | Expected cell counts $\geq 5$; independence |
| Linear regression | Linearity; normal residuals; constant variance |
| Logistic regression | Independence; no complete separation |
| Wilcoxon rank-sum | Independence; similar distribution shape |
:::

. . .

**How to check:** histograms, Q-Q plots, `plot(lm(...))`, raw contingency table

---

## Effect Size, Not Just p-Values

**p-value:** Is this result likely due to chance?

**Effect size:** How *big* is this effect? Is it clinically meaningful?

. . .

| Test | Effect Size | Benchmark |
|------|------------|-----------|
| t-test | Cohen's $d$ | 0.2 small, 0.5 medium, 0.8 large |
| Linear regression | $R^2$ | Proportion of variance explained |
| Logistic regression | Odds Ratio | Odds multiplier per unit change |
| Chi-square | Cramér's $V$ | 0-1 scale of association strength |

. . .

**Clinical example:** Drug lowers SBP by 2 mmHg, p = 0.001 — significant but not clinically meaningful.

. . .

Always report both the p-value and the effect size.
