---
title: "Module 5 — Case Study: Putting It All Together"
subtitle: "Introduction to Statistics for Health Science"
author: "Josh Lim"
date: today
format: revealjs
execute:
  echo: true
---

```{r}
#| echo: false
nhanes <- read.csv("../../data/nhanes_sample.csv", header = TRUE,
                   stringsAsFactors = FALSE)
```

# The Scenario

## Your PI's Research Questions

Your PI hands you the NHANES dataset and says:

. . .

**Q1:** "Is BMI associated with systolic blood pressure?"

. . .

**Q2:** "Is smoking associated with hypertension?"

. . .

**Q3:** "After adjusting for BMI and physical activity, is smoking independently associated with diabetes?"

. . .

**Your job:** The analysis — from raw data to written interpretation — is entirely up to you.

# Step 1: Data Cleaning

## Step 1: Create Derived Variables

```{r}
#| echo: true
nhanes_clean <- nhanes

nhanes_clean$hypertensive <- ifelse(nhanes_clean$SBP >= 140, 1, 0)
nhanes_clean$obese        <- ifelse(nhanes_clean$BMI >= 30,  1, 0)
```

# Step 2: Baseline Description

## Step 2: Who Is in the Dataset? {.smaller}

```{r}
#| echo: true
nrow(nhanes_clean)
table(nhanes_clean$Smoker)
```

. . .

```{r}
#| echo: true
mean(nhanes_clean$BMI[nhanes_clean$Smoker == 0])  # non-smokers
mean(nhanes_clean$BMI[nhanes_clean$Smoker == 1])  # smokers
mean(nhanes_clean$SBP[nhanes_clean$Smoker == 0])
mean(nhanes_clean$SBP[nhanes_clean$Smoker == 1])
```

. . .

```{r}
#| echo: true
sex_tab <- table(nhanes_clean$Smoker, nhanes_clean$Sex)
prop.table(sex_tab, margin = 1)
```

---

## What Table 1 Tells You

- Are your groups comparable at baseline?
- Imbalances may indicate confounding
- Guides which variables to include as covariates

# Q1: BMI → Systolic Blood Pressure

## Q1 Overview

- **Outcome:** SBP (continuous)
- **Predictor:** BMI (continuous)
- **Test:** Simple linear regression — `lm()`

---

## Q1: Summarize

```{r}
#| echo: true
mean(nhanes_clean$SBP);  sd(nhanes_clean$SBP)
mean(nhanes_clean$BMI);  sd(nhanes_clean$BMI)
```

. . .

```{r}
#| echo: true
# Pearson correlation: direction and strength before modeling
cor(nhanes_clean$BMI, nhanes_clean$SBP)
```

---

## Q1: Visualize

```{r}
#| echo: true
#| fig-height: 5
#| fig-width: 6
#| output-location: column-fragment
plot(nhanes_clean$BMI, nhanes_clean$SBP,
     main = "BMI vs Systolic Blood Pressure",
     xlab = "BMI (kg/m²)",
     ylab = "SBP (mmHg)",
     pch  = 16,
     col  = rgb(0, 0, 1, 0.3))
```

A positive trend? Now let's quantify it.

---

## Q1: Model

```{r}
#| echo: true
#| output-location: column-fragment
fit_lm <- lm(SBP ~ BMI, data = nhanes_clean)
summary(fit_lm)
```

---

## Q1: Model — Regression Line

```{r}
#| echo: true
#| fig-height: 5
#| fig-width: 6
#| output-location: column-fragment
plot(nhanes_clean$BMI, nhanes_clean$SBP,
     main = "BMI Predicts SBP",
     xlab = "BMI (kg/m²)", ylab = "SBP (mmHg)",
     pch  = 16, col = rgb(0, 0, 0.8, 0.3))
abline(fit_lm, col = "red", lwd = 2)
```

---

## Q1: Assumption Check

```{r}
#| echo: true
#| fig-height: 5
#| fig-width: 9
par(mfrow = c(2, 2))
plot(fit_lm)
par(mfrow = c(1, 1))
```

---

## Q1: Interpret

- **Slope:** `r round(coef(fit_lm)[2], 2)` mmHg increase in SBP per 1 kg/m² increase in BMI

. . .

- **R²:** BMI explains `r round(summary(fit_lm)$r.squared * 100, 1)`% of the variability in SBP

. . .

> "Each 1-unit increase in BMI was associated with a `r round(coef(fit_lm)[2], 2)` mmHg increase in SBP (95% CI: `r round(confint(fit_lm)[2,1], 2)` to `r round(confint(fit_lm)[2,2], 2)`; p < 0.05)."

# Q2: Smoking → Hypertension

## Q2 Overview

- **Outcome:** Hypertensive (binary: 0/1)
- **Predictor:** Smoker (binary: 0/1)
- **Test:** Chi-square + `fisher.test()` for the Odds Ratio

---

## Q2: Summarize

```{r}
#| echo: true
ct <- table(Hypertensive = nhanes_clean$hypertensive,
            Smoker       = nhanes_clean$Smoker)
dimnames(ct)$Hypertensive <- c("No", "Yes")
dimnames(ct)$Smoker       <- c("Non-smoker", "Smoker")
addmargins(ct)
```

---

## Q2: Visualize

```{r}
#| echo: true
#| fig-height: 5
#| fig-width: 6
#| output-location: column-fragment
props <- prop.table(ct, margin = 2)
barplot(props["Yes", ] * 100,
        names.arg = c("Non-smoker", "Smoker"),
        ylim   = c(0, 100),
        main   = "Hypertension Prevalence\nby Smoking Status",
        ylab   = "% Hypertensive",
        col    = c("lightblue", "salmon"),
        border = "white")
```

---

## Q2: Model

```{r}
#| echo: true
chisq.test(ct)
```

. . .

```{r}
#| echo: true
ft <- fisher.test(ct)
ft$estimate   # Odds Ratio
ft$conf.int   # 95% CI
ft$p.value    # exact p-value
```

---

## Q2: Interpret

> "Smokers had `r round(ft$estimate, 2)` times the odds of hypertension compared to non-smokers (95% CI: `r round(ft$conf.int[1], 2)`–`r round(ft$conf.int[2], 2)`; p = `r round(ft$p.value, 3)`)."

. . .

- OR > 1 → smoking associated with higher odds of hypertension
- Does the CI include 1? If yes → not significant at α = 0.05

# Q3: Smoking → Diabetes (Adjusted)

## Q3 Overview

- **Outcome:** Diabetes (binary: 0/1)
- **Predictors:** Smoker + BMI + PhysActive
- **Test:** Multivariable logistic regression — `glm(..., family = binomial)`

. . .

Why adjust? BMI and physical activity are plausible confounders — we want the smoking effect *holding those constant*.

---

## Q3: Summarize

```{r}
#| echo: true
# Diabetes prevalence by smoking status
mean(nhanes_clean$Diabetes[nhanes_clean$Smoker == 0])  # non-smokers
mean(nhanes_clean$Diabetes[nhanes_clean$Smoker == 1])  # smokers
```

. . .

```{r}
#| echo: true
# Mean BMI by diabetes status
mean(nhanes_clean$BMI[nhanes_clean$Diabetes == 0])
mean(nhanes_clean$BMI[nhanes_clean$Diabetes == 1])
```

---

## Q3: Visualize

```{r}
#| echo: true
#| fig-height: 5
#| fig-width: 9
#| output-location: column-fragment
par(mfrow = c(1, 2))

diab_props <- prop.table(
  table(nhanes_clean$Smoker, nhanes_clean$Diabetes),
  margin = 1)
barplot(diab_props[, 2] * 100,
        names.arg = c("Non-smoker", "Smoker"),
        ylim = c(0, 30),
        main = "Diabetes Prevalence\nby Smoking Status",
        ylab = "% with Diabetes",
        col  = c("lightblue", "salmon"), border = "white")

boxplot(BMI ~ Diabetes, data = nhanes_clean,
        names = c("No Diabetes", "Diabetes"),
        main  = "BMI by Diabetes Status",
        ylab  = "BMI (kg/m²)",
        col   = c("lightblue", "salmon"))

par(mfrow = c(1, 1))
```

---

## Q3: Model

```{r}
#| echo: true
#| output-location: column-fragment
fit_log <- glm(Diabetes ~ Smoker + BMI + PhysActive,
               data   = nhanes_clean,
               family = binomial)
summary(fit_log)
```

---

## Q3: Odds Ratios

```{r}
#| echo: true
exp(coef(fit_log))
```

. . .

```{r}
#| echo: true
exp(confint(fit_log))
```

. . .

- **Smoker OR > 1:** higher odds of diabetes, adjusting for BMI and activity
- **PhysActive OR < 1:** physically active participants have lower odds
- **CIs excluding 1:** statistically significant predictors

---

## Q3: Interpret — Forest Plot

```{r}
#| echo: true
#| fig-height: 5
#| fig-width: 6
#| output-location: column-fragment
ors  <- exp(coef(fit_log))[-1]
cis  <- exp(confint(fit_log))[-1, ]
labs <- c("Smoker", "BMI", "Phys. Active")
plot(ors, seq_along(ors),
     xlim = c(0.5, max(cis) + 0.2),
     ylim = c(0.5, length(ors) + 0.5),
     pch = 15, cex = 1.5,
     xlab = "Odds Ratio (95% CI)",
     yaxt = "n", ylab = "",
     main = "Predictors of Diabetes")
arrows(cis[, 1], seq_along(ors), cis[, 2], seq_along(ors),
       angle = 90, code = 3, length = 0.05, col = "black")
abline(v = 1, lty = 2, col = "red")
axis(2, at = seq_along(ors), labels = labs, las = 1)
```

---

## Q3: Interpret — Plain English

> "After adjusting for BMI and physical activity, smoking remained a significant predictor of diabetes (OR = `r round(exp(coef(fit_log)["Smoker"]), 2)`, 95% CI: `r round(exp(confint(fit_log))["Smoker",1], 2)`–`r round(exp(confint(fit_log))["Smoker",2], 2)`)."

. . .

> "Each 1-unit increase in BMI was associated with higher odds of diabetes (OR = `r round(exp(coef(fit_log)["BMI"]), 2)`) after adjusting for smoking and physical activity."

# Limitations

## What Can't We Conclude?

**Causation vs. Association:**

- BMI is associated with higher SBP — this does NOT mean BMI causes it
- Cross-sectional data: no temporal ordering

. . .

**Confounding:**

- A third variable (e.g., Age, diet) could explain the association
- Logistic regression adjusts for measured confounders only

. . .

**Study design limitations:**

- NHANES is a survey — sampling weights needed for representative estimates
- Self-reported smoking may have measurement error

# Writing Up Results

## Results Paragraph Template

> "[Outcome] was [significantly / not significantly] associated with [predictor] ([test name], [statistic] = [value], p = [value]). [Direction and magnitude, with units.] The 95% CI for [effect measure] was [lower] to [upper]."

. . .

**Your turn — 5 minutes:**

Pick any one finding and write one sentence. Include:

1. The two variables involved
2. The direction of the association
3. The p-value or OR
4. What it means clinically

. . .

Share with your neighbor. Would a clinician understand it?

## Thank You!

**What you covered today:**

- Module 1: Data science primer — loading, subsetting, wrangling
- Module 2: Descriptive statistics and visualization
- Module 3: Inferential statistics fundamentals
- Module 4: Choosing the right statistical test
- Module 5: End-to-end analysis pipeline


## Thank You!

**Workshop materials:**

- Scripts: [github.com/joshlim789/JCSU_R_Workshop](https://github.com/joshlim789/JCSU_R_Workshop)
- Dataset: NHANES public use files at cdc.gov/nchs/nhanes
- Contact: josh.lim@duke.edu

. . .

<br>

**Good luck with your research!**
