---
title: "Module 1 — Data Science Primer"
subtitle: "Introduction to Statistics for Health Science"
author: "Josh Lim"
date: today
format: revealjs
execute:
  echo: true
---

# Part A — Where Does Health Data Come From?

## Sources of Health Data

Health research runs on data — where does it come from?

. . .

- **EHR** — diagnoses, labs, prescriptions from clinical care
- **National Surveys** — NHANES, BRFSS; US population samples
- **Clinical Trials** — participants assigned to treatment or control
- **Disease Registries** — cancer, trauma; track conditions over time
- **Wearables & Apps** — Fitbit, Apple Watch; passive physiological data

. . .

**Our data today:** NHANES — nationally representative survey

---

## Tidy Data: The Gold Standard

For analysis, data should be **tidy**:

> Each row = one **observation**. Each column = one **variable**.

. . .

**Tidy (ready to analyze):**

| ID  | Age | BMI  | Smoker |
|-----|-----|------|--------|
| 101 | 45  | 27.3 | 1      |
| 102 | 62  | 31.0 | 0      |
| 103 | 33  | 24.8 | 1      |

---

## Messy Data (Common in Practice)

| Variable | Person_1 | Person_2 | Person_3 |
|----------|----------|----------|----------|
| Age      | 45       | 62       | 33       |
| BMI      | 27.3     | 31.0     | 24.8     |
| Smoker   | 1        | 0        | 1        |

. . .

- Variables as rows, observations as columns
- Must be reshaped before analysis

# Part B — Reading Data In

## Loading a CSV with `read.csv()`

The most common data format in health research: `.csv`

```{r}
#| echo: true
# read.csv() loads a CSV file into a data frame
nhanes <- read.csv("../../data/nhanes_sample.csv", header = TRUE)
```

. . .

**Key arguments:**

| Argument | Purpose | Default |
|----------|---------|---------|
| `file` | path to the file | required |
| `header` | first row = names? | `TRUE` |
| `sep` | field separator | `","` |
| `na.strings` | what counts as NA? | `"NA"` |

---

## After Loading: The Data Frame

- `nhanes` is now a **data frame** — the workhorse object in R
- Rows = observations (patients)
- Columns = variables

. . .

Always inspect immediately after loading:

- Check dimensions
- Check variable names
- Look at the first few rows

---

## First Look: Dimensions and Names

```{r}
#| echo: true
# nrow() counts observations; ncol() counts variables
nrow(nhanes)
ncol(nhanes)

# names() shows all column names
names(nhanes)
```

---

## First Look: Peeking at the Data

```{r}
#| echo: true
# head() shows the first 6 rows by default
head(nhanes)
```


---

## `str()` — Structure at a Glance

`str()` is your best first-look tool — type + preview for every column.

```{r}
#| echo: true
# str() prints structure: obs count, variables, type, first values
str(nhanes)
```

# Part C — Subsetting and Summarizing

## Accessing Columns

Two equivalent ways to pull out a single column:

```{r}
#| echo: true
#| eval: false
# Dollar sign notation: most common
nhanes$BMI

# Bracket notation with column name
nhanes[, "BMI"]
```

. . .

- Both produce the **same vector**
- Use `$` for interactive work
- `[,]` is useful inside functions and loops

---

## Subsetting Rows with `[rows, cols]`

We often want to analyze a specific group — like smokers only.

```{r}
#| echo: true
# nhanes[row_condition, "column_name"]
# Leave column blank to keep all columns
smokers <- nhanes[nhanes$Smoker == 1, ]
nrow(smokers)
```

. . .

```{r}
#| echo: true
# Multiple conditions with & (AND)
older_smokers <- nhanes[nhanes$Smoker == 1 & nhanes$Age > 40, ]
nrow(older_smokers)
```

. . .

Note: `==` is *comparison*; `=` is *assignment*. Easy bug to make!

---

## Creating New Variables with `ifelse()`

We can derive new variables from existing ones — very common in health research.

```{r}
#| echo: true
# ifelse(test, value_if_TRUE, value_if_FALSE)
nhanes$obese <- ifelse(nhanes$BMI >= 30, 1, 0)
table(nhanes$obese)
```

. . .

```{r}
#| echo: true
# Hypertension: SBP >= 140 mmHg is a common clinical cutoff
nhanes$hypertensive <- ifelse(nhanes$SBP >= 140, 1, 0)
table(nhanes$obese, nhanes$hypertensive)
```

---

## `summary()` — Quick Statistical Overview

```{r}
#| echo: true
# Gives min, Q1, median, mean, Q3, max for numeric columns
summary(nhanes)
```

. . .

- Fast sanity check for the whole dataset
- Spot impossible values immediately

---

## `table()` — Counting Categories

```{r}
#| echo: true
# One-way table: how many males vs. females?
table(nhanes$Sex)
```

. . .

```{r}
#| echo: true
# Two-way table: rows = Sex, columns = Smoker
table(nhanes$Sex, nhanes$Smoker)
```

. . .

Two-way tables are the foundation of chi-square tests and odds ratios (Module 03).

---

## Module 1 Recap

**What you can now do:**

- Load CSV data with `read.csv()`
- Inspect a dataset with `str()`, `head()`, `dim()`
- Subset rows with `subset()` and filter with conditions
- Create new variables with `ifelse()`
- Summarize with `summary()`, `table()`, and `prop.table()`

---

## What's Next

**Module 2 — Descriptive Statistics**

- Measures of center: mean and median
- Measures of spread: SD, IQR, range
- Data types and which summaries apply to each
- Visualization: histograms, boxplots, scatter plots, bar charts
